\Rpackage{MSnbase}: labelled MS2 data pre-processing, visualisation and quantification.

### 2011-09-08 inst/doc/MSnbase-demo/MSnbase-demo.Rnw
new io functions and doc
 o added a readMSnSet function <2011-09-08 Thu>
 o added a write.MSnSet("MSnSet") method  <2011-09-08 Thu>
 o vignette/man updates - mainly data import section documenting 
   readMSData and readMgfData <2011-09-08 Thu>



git-svn-id: file:///home/git/hedgehog.fhcrc.org/bioconductor/trunk/madman/Rpacks/MSnbase@57904 bc3139a8-67e5-0310-9ffc-ced21a209358

\newcommand{\R}{\texttt{R} }
\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
%% colors
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}

@@

\subsection{Importing experiments}\label{sec:msnexp}

@@

\Rpackage{MSnbase} imports raw MS data stored in \texttt{mzXML} 
format \citep{Pedrioli2004} with the function \Rfunction{readMzXMLData} function. 
See \Rfunction{?readMzXMLData} for more details.
More formats, notably \texttt{mzML} \citep{Martens2010}, will be added at a later stage. 
Either MS1 or MS2 spectra can be loaded at a time by setting the \texttt{msLevel} parameter accordingly.
In this document, we will use a slightly bigger data set, \Robject{itraqdata}, provided with 
\Rpackage{MSnbase}. It includes feature metadata, accessible with the \Rfunction{fData} method. 
The metadata includes identification data for the \Sexpr{length(itraqdata)} MS2 spectra.

@@

\subsection{Purity correction}
\label{sec:purcor}

Impurities in the reporter reagents can bias the results and can be corrected when 
manufacturers provide correction coefficients. 
These generally come as percentages of each reporter ion that have masses differing 
by -2, -1, +1 and +2 Da from the nominal reporter ion mass due to isotopic variants.  
The \Rfunction{purityCorrect} method applies such correction to \Robject{MSnSet} 
instances. It also requires a square matrix as second argument, \Robject{impurities}, 
that defines the relative percentage of reporter in the quantified each peak. 
See \texttt{?purityCorrect} for more details.

>>>>

\newcommand{\R}{\texttt{R} }
\newcommand{\Rfunction}[1]{{\texttt{#1}}}
\newcommand{\Robject}[1]{{\texttt{#1}}}
%%\newcommand{\Rpackage}[1]{{\textit{#1}}}
\newcommand{\Rpackage}[1]{{\mbox{\normalfont\textsf{#1}}}}
\newcommand{\email}[1]{\href{mailto:#1}{\normalfont\texttt{#1}}}
%% colors
\definecolor{Red}{rgb}{0.7,0,0}
\definecolor{Blue}{rgb}{0,0,0.8}

@@

\subsection{Importing experiments}\label{sec:io}

\Rpackage{MSnbase} is able to import raw MS data stored in one of the \texttt{XML}-based formats 
as well as peak lists in the \texttt{mfg} format\footnote{%%
Mascot Generic Format -- \url{http://www.matrixscience.com/help/data\_file\_help.html\#GEN}}

\paragraph{Raw data} The \texttt{XML}-based formats, \texttt{mzXML} \citep{Pedrioli2004}, 
\texttt{mzData} \citep{Orchard2007} and \texttt{mzML} \citep{Martens2010} can be imported 
with the \Rfunction{readMSData} function, as illstrated below 
(see \Rfunction{?readMSData} for more details).

<<readdata, echo=TRUE, cache=FALSE>>=
file <- dir(system.file(package="MSnbase",dir="extdata"),
            full.name=TRUE,
            pattern="mzXML$")
rawdata <- readMSData(file, msLevel=2, verbose=FALSE)
@

Either MS1 or MS2 spectra can be loaded at a time by setting the \texttt{msLevel} parameter accordingly.
In this document, we will use the \Robject{itraqdata} data set, provided with 
\Rpackage{MSnbase}. It includes feature metadata, accessible with the \Rfunction{fData} accessor. 
The metadata includes identification data for the \Sexpr{length(itraqdata)} MS2 spectra.

\paragraph{Peak lists} Peak lists can often be exported after spectrum processing 
from vendor-specific software and are also used as input to search engines. 
Peak lists in \texttt{mgf} format can be imported with the function 
\Rfunction{readMgfData} (see \Rfunction{?readMgfData} for details)
to create experiment objects. 
Experiments or individual spectra can be exported to an \texttt{mgf} file with 
the \Rfunction{writeMgfData} methods (see \Rfunction{?writeMgfData} for details and examples).

See also section \ref{sec:io2} to import quantitative data stored in spreadsheets into \R 
for further processing using \Rpackage{MSnbase}.

\subsection{MS experiments}\label{sec:msnexp}

@@

@@

\subsection{Importing quantitation data}\label{sec:io2}

If quantitation data is already available as a spreadsheet, it can be 
imported, along with additional optional feature and sample (pheno) meta data, 
with the \Rfunction{readMSnSet} function. This function takes the 
respective text-based spreadsheet (comma- or tab-separated) file names 
as argument to create a valid \Robject{MSnSet} instance.

Note that the quantitation data of \Robject{MSnSet} objects can also be 
exported to a text-based spreadsheet file using the \Rfunction{write.exps} 
method.

\subsection{Peak adjustments}\label{sec:purcor}

\paragraph{Single peak adjustment} In certain cases, peak intensities need to be adjusted 
as a result of peak interferance. For example, the $+1$ peak of the phenylalanine (F, Phe) 
immonium ion (with m/z 120.03) inteferes with the 121.1 TMT reporter ion. 
Below, we calculate the relative intensity of the +1 peaks compared to the main peak using the 
\Rpackage{Rdispo} package.

<<pheplus1, echo=TRUE, cache=FALSE>>=
library(Rdisop)
## Phenylalanine immonium ion
Fim <- getMolecule("C8H10N")
getMass(Fim)
isotopes <- getIsotope(Fim)
F1 <- isotopes[2,2]
F1
@

If desired, one can thus specifically quantify the F immonium ion in the MS2 spectrum, 
estimate the intensity of the +1 ion (\Sexpr{round(F1,4)}\% of the F peak) and substract 
this calculated value from the 121.1 TMT reporter intensity.

The above principle can also be generalised for a set of overlapping peaks, 
as described below.

\paragraph{Reporter ions purity correction} Impurities in the reporter reagents can also 
bias the results and can be corrected when manufacturers provide correction coefficients. 
These generally come as percentages of each reporter ion that have masses differing 
by -2, -1, +1 and +2 Da from the nominal reporter ion mass due to isotopic variants.  
The \Rfunction{purityCorrect} method applies such correction to \Robject{MSnSet} 
instances. It also requires a square matrix as second argument, \Robject{impurities}, 
that defines the relative percentage of reporter in the quantified each peak. 
See \texttt{?purityCorrect} for more details.
